---
title: "Exercise 3, Data analysis"
author: "Pauliina Anttila"
date: "17 11 2017"
output: html_document
---

# 3. Logistic regression

## Description of the dataset

**This week we are working with a joined dataset that contains two portuguese student alcohol consumption data sets. Originally, the data sets are found from the UCI Machine Learning Repository under the name "Student performance data set" (donated Nov 2014). The original data sets contains data about student performance in secondary education of two Portuguese schools in two distinct subjects: mathematics and Portuguese language. The data was collected by using school reports and questionnaires. We wrangled the data so that we combined the data sets (mathematics and Portuguese language).**

The wrangled data set contains 382 observations of 35 variables. The variables include the following:

1 school - student's school

2 sex - student's sex

3 age - student's age

4 address - student's home address type

5 famsize - family size

6 Pstatus - parent's cohabitation status

7 Medu - mother's education

8 Fedu - father's education

9 Mjob - mother's job

10 Fjob - father's job

11 reason - reason to choose this school

12 guardian - student's guardian

13 traveltime - home to school travel time

14 studytime - weekly study time 

15 failures - number of past class failures 

16 schoolsup - extra educational support

17 famsup - family educational support

18 paid - extra paid classes within the course subject (Math or Portuguese)

19 activities - extra-curricular activities

20 nursery - attended nursery school

21 higher - wants to take higher education

22 internet - Internet access at home

23 romantic - with a romantic relationship

24 famrel - quality of family relationships 

25 freetime - free time after school 

26 goout - going out with friends

27 Dalc - workday alcohol consumption

28 Walc - weekend alcohol consumption

29 health - current health status 

30 absences - number of school absences 

31 G1 - first period grade

32 G2 - second period grade 

33 G3 - final grade 

34 alc_use - the average of alcohol consumption on weekdays and weekends

35 high_use - TRUE if "alc_use is higher than 2"

```{r}
alc <- read.csv (file = "~/Documents/GitHub/IODS-project/data/alc.csv", header = TRUE, sep = ",", dec = ".") # reading data into R
library(dplyr)
colnames(alc) # printing out the names of the variables
```

```{r}
dim(alc) # there are 382 observations of 35 variables in the data
```

## Exploring the data set

### My hypotheses:

*1. Older students consume high amounts of alcohol more often than younger ones.*

*2. Those who have high use of alcohol tend to fail more often.*

*3. Those who have extra-curricular activities consume less alcohol.*

*4. Those who want to take higher education consume less alcohol.*

**1. Age and alcohol consumption:**

```{r}
library(ggplot2)
p1 <- ggplot(alc, aes(x = high_use, y = age))
p1 + geom_boxplot() + ylab("age")
```


```{r}
t1 <- table(alc$high_use, alc$age) # crosstabulation of "high_use" and "age"
t1
```

**According to the boxplots and crosstabulation, it seems that older students consume high amounts of alcohol more often than the younger ones.**

**2. Alcohol consumption and failures:**

```{r}
t2 <- table(alc$high_use, alc$failures) # crosstabulation of "high_use" and "failures"
t2
```

```{r}
counts <- table(alc$high_use, alc$failures)
barplot(counts, main="High alcohol use and failures",
  xlab="Failures",
 	legend = rownames(counts), beside=TRUE)
```

Number of failures in the x-axis and count of students on the y-axis. Grouped by "high_use" (either TRUE or FALSE).

**Only a small proportion of students have failed and the absolute counts of failures are fairly similar in both groups (those who use high amounts of alcohol vs those who do not). It is to be noted, that there is a lot more students in the latter group (268 vs 114).**

**3. Alcohol consumption and activities:**

```{r}
t3 <- table(alc$high_use, alc$activities)
t3
```
```{r}
counts <- table(alc$high_use, alc$activities)
barplot(counts, main="High alcohol use and activities",
  xlab="Activities",
 	legend = rownames(counts), beside=TRUE)
```

**The proportion of students who consume high amounts of alcohol is bigger in the group that has no activities.**

**4. Alcohol consumption and the urge to take higher education:**

```{r}
t4 <- table(alc$high_use, alc$higher)
t4
```
```{r}
counts <- table(alc$high_use, alc$higher)
barplot(counts, main="High alcohol use and the urge to take higher education",
  xlab="The urge to take higher education",
 	legend = rownames(counts), beside=TRUE)
```

**The majority of the students want to take higher education.**

## Analysis with logistic regression

```{r}
my_model <- glm (formula = high_use ~ age + failures + activities + higher, data = alc, family ="binomial") # Logistic regression model with high_use as target variable and age, failures, activities and higher education target as explanatory variables.
summary(my_model)
```

```{r}
coef(my_model) # coefficients of the model
```

```{r}
OR <- coef(my_model) %>% exp # computing odds ratios
CI <- confint(my_model) %>% exp # computing confidence intervals
cbind(OR,CI) # printing ORs and CIs
```

**The only examined variable that is statistically significantly associated to high use of alcohol is failures. The odds ratio of failures is 1.55 with confidence interval of 1.06-2.29 (p-value 0.024). With high consumption of alcohol the odds of failing is 1.55 times higher than with smaller alcohol consumption.**

**The previous data exploring and logistic regression model show similar trends with other variables, but they are not statistically significant. **

## Exploring the predictive power of my model

```{r}
my_model2 <- glm(high_use ~ failures, data = alc, family = "binomial") # fitting the model with the only variable (failures) that had statistically significant relationship with high use of alcohol
probabilities <- predict(my_model2, type = "response") # predicting the probability of high use
alc <- mutate(alc, probability = probabilities) # adding the predicted probabilities to 'alc'
alc <- mutate(alc, prediction = probability > 0.5) # using the probabilities to make a prediction of high_use
table(high_use = alc$high_use, prediction = alc$prediction) # tabulation of the target variable versus the predictions
```

```{r}
library(dplyr)
library(ggplot2)
g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction)) # initializing a plot of 'high_use' versus 'probability' in 'alc'
g + geom_point() # defining the geom as points and drawing the plot
```

```{r}
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table() %>% addmargins() # tabulating the target variable versus the predictions
```

```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong) 
} # defining a loss function (mean prediction error)
loss_func(class = alc$high_use, prob = alc$probability) # calling loss_func to compute the average number of wrong predictions in the (training) data
```

**The average amount of wrong predictions in the (training) data is 29,8%. The proportion is quite high but still better than pure quessing.**


```{r}

```


```{r}

```


```{r}

```

```{r}

```

```{r}

```