---
title: "Exercise 5 data analysis"
author: "Pauliina Anttila"
date: "5 12 2017"
output: html_document
---
# 5. Dimensionality reduction techniques

**This week we are working with the 'human' data. Before analysis, we wrangled the data.**

**You can see data wrangling part here:** https://github.com/pauliinaanttila/IODS-project/blob/master/data/create_human.R. 

**Here you can see the original data from the United Nations Development Programme:** http://hdr.undp.org/en/content/human-development-index-hdi.

## Exploring the data:

```{r}
human <-read.csv(file = "~/Documents/GitHub/IODS-project/data/human.csv", header = TRUE, sep = ",", dec = ".", row.names=1) # reading the data into R
```

```{r}
dim(human) # exporing the dimensions of the data
```

```{r}
str(human) # eploring the structure of the data
```

**The dataset contains 155 observations of 8 variables. The data combines several indicators from most countries in the world. The variables include the following:**

"Country" = Country name (as row names in the data)

**About health and knowledge:**

"GNI" = Gross National Income per capita

"Life.Exp" = Life expectancy at birth

"Edu.Exp" = Expected years of schooling 

"Mat.Mor" = Maternal mortality ratio

"Ado.Birth" = Adolescent birth rate

**Empowerment:**

"Parli.F" = Percetange of female representatives in parliament

"Edu2.F" = Proportion of females with at least secondary education

"Edu2.M" = Proportion of males with at least secondary education

"Labo.F" = Proportion of females in the labour force

"Labo.M" " Proportion of males in the labour force

"Edu2.FM" = Edu2.F / Edu2.M

"Labo.FM" = Labo2.F / Labo2.M

## Overview of the data:

```{r}
library(GGally)
library(ggplot2)
ggpairs(human, lower = list(combo = wrap("facethist", bins = 20)))
```

**Expected years of schooling seem to been pretty normally distributed. Otherwise the distributions have considerable skewness, a positive one for Gross National Income per capita, maternal mortality ratio, adolescent birth rate and percetange of female representatives in parliament, and a negative one for the rest.**

**It seems that maternal mortality and adolescent births have a positive correlation. Also life expectancy at birth and expected years of schooling have a positive correlation.**

```{r}
summary(human)
```

**Life expectancy at birth varies between 49.0 - 83.5 years with the median of 74.2. Expected years of schooling vary between 5.4 - 20.2 years with the median of 13.5. The maternal mortality ratio varies hugely, it ranges from 1.0 - 1100.0.**

## Principal component analysis (PCA):

**Now we are going to perform principal component analysis (PCA) on the not standardized 'human' data. It is statistical method used in dimension reduction.**

```{r}
pca_human <- prcomp(human)
summary(pca_human)
```

**Above in the summary is shown the variability captured by the principal components.**

```{r}
biplot(pca_human, choices = 1:2, cex = c(0.8,1), col = c("grey40", "deeppink2")) # drawing a biplot of the principal component representation and the original variables
```

**A biplot displaying the observations by the first two principal components, along with arrows representing the original variables.**

**Before standardizing the dataset, the variances are so different between the variables that it is difficult to interpret the PCA in a sensible manner. The GNI has more leverage in the unstandardized data because it has a huge variation in values.**

##Standardizing the dataset and performing PCA again:

```{r}
human_std <- scale(human) # standardizing the dataset
pca_human_std <- prcomp(human_std)
summary(pca_human_std)
```

**After standardization, the variances of the variables are more alike.**

```{r}
biplot(pca_human_std, choices = 1:2, cex = c(0.8,1), col = c("grey40", "deeppink2"))
```

**Now we can truly see the connections between variables.**

**It seems that maternal mortality ratio and adolescent birth rate have a high positive correlation (small angle between the arrows).**

**Percentage of female representatives in parliament is highly positively correlated to the ratio of proportion of females vs males in the labour force.**

**Life expectancy at birth, expected years of schooling and GNI are positively correlated.**

# Interpretation of the first two principal component dimensions:

**Factors contributing to PC1 are maternal mortality ratio and adolescent birth rate (arrows to the same direction).**

**Variables contributing to PC2 are Parli.F and Labo.FM.**

# Working with 'tea' dataset:

```{r}
library(FactoMineR)
data(tea) # loading the 'tea' dataset from the package FactoMineR
dim(tea)
```

```{r}
str(tea)
```

```{r}
colnames(tea)
```

**The 'tea' dataset has 300 observations of 36 variables concerning tea consumption (for example if tea is consumed during breakfast, lunch, dinner etc.).**

**Let's create a bit smaller dataset:**

```{r}
keep_columns <- c("Tea", "How", "how", "sugar", "where", "resto", "pub")
tea_selected <- tea[, keep_columns] # creating a smaller dataset with the selected columns

```{r}
dim(tea_selected)
```

**The new 'tea_selected' dataset contains 300 observations of 7 variables. Summary of the variables is seen below:**

```{r}
summary(tea_selected)
```

```{r}
library(dplyr)
library(tidyr)
gather(tea_selected) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() 
```

### Multiple correspondence analysis with the 'tea' dataset:

```{r}
mca <- MCA(tea_selected, graph = FALSE)
summary(mca)
```

```{r}
mca <- MCA(tea_selected, graph = TRUE)
```
```{r}
plot(mca, invisible=c(), habillage = "quali")
```